### 2023-06-16 manuscript/appendices/B_zcurve.qmd
dkfjwen fhjk !!!!!

::: content-hidden
{{< include ../macros.tex >}}
\def\C{\mathcal{C}}
:::

# On the quasi-bijectivity and continuity of the Z-mapping

We study the Z-curve $Z^{-1}$

## The Z-curve is not injective {#sec-zcurve-noninjective}

this is common misconception, and people write it is a bijection, when really it is not

The Z-curve : [0,1] -> [0,1]^2 is not injective since the binary expansion is ambiguous

x = 0.10000... = 0.011111...

=> alpha = 0.1y0y0y0  !=   alpha = 0.0y1y1y1y1

but the point on which the expansion is ambiguous form a set of zero measure, so the integral is still ok

on the set [0,1] \ {problems}, Z-curve is bijective

injectivity is the property that f(x) != f(y)  iff  x != y,  (equiv f(x) == f(y) iff x == y)

Z^{-1} defines a space-filling curve

not a problem because we are just repeating the same population twice

## $Z^{-1}$ defines a space-filling curve on a subset of $[0,1]$ {#sec-zcurve-lebesgue}

@@

talk about the fact we have other space filling curves, such as hilbert curve or peano curve, that are continuous and surjective [0,1] -> [0,1]^2. we use the Z-curve because it's easy to implement numerically

>>>>

::: {.content-hidden}
{{< include ../macros.tex >}}
\def\C{\mathcal{C}}
\def\P{\mathcal{P}}
\def\B{\mathcal{B}}
:::

# The Z-curve is not injective {#sec-zcurve-noninjective}

Let us study, in a bit more detail, the properties of the Z-curve $Z^{-1}$ introduced in @sec-local, and, by extension, the properties of the Z-mapping $Z$.

## Ambiguity of binary expansions

Although it might seem that the Z-curve defines a bijection (this is often falsely relayed in online resources, and was even initially believed by Cantor when he discovered this function, before being corrected by Dedekind @fleron1994cantorhistory), this is not the case.

The problem with the definition is the ambiguity of (binary) expansions. If we allow infinite sum expansions, then the two representations of $\tfrac 12$ are equivalent:

$$
\begin{aligned}
\frac 12 &= 0.1000 \cdots = \sum_{l=1}^\infty b_l 2^{-l},\ \text{where}\ b_1=1, b_{l>1}=0 \\
&= 0.0111 \cdots = \sum_{l=1}^\infty b^\prime_l 2^{-l},\ \text{where}\ b^\prime_1=0, b^\prime_{l>1}=1
\end{aligned}
$$

Although $Z^n$ are still bijections, this ambiguity is problematic for the $n\to\infty$ limit, since we can find multiple pre-images for the same $\alpha$ mapped by $Z$. This illustrate this, let us take a finite integer $m$, and write a counter-example of bijectivity. Similarly to the main text, we define the binary expansions of $v_1$ and $v_2$:

$$
\begin{aligned}
v_1 &= \sum_{l=1}^\infty b^1_l 2^{-l} = 0.b^1_1b^1_2 \cdots b^1_{m} b^1_{m+1} \cdots \\
v_2 &= \sum_{l=1}^\infty b^2_l 2^{-l} = 0.b^2_1b^2_2 \cdots b^2_{m} b^2_{m+1} \cdots
\end{aligned}
$$

We now consider the $l > m$ binary digits of $v_1$ to give ambiguous representations of $v_1^{(m)} + 2^{-m-1}$, where $v_1^{(m)}$ is the finite-$m$ truncation of $v_1$.

$$
\begin{aligned}
v_1 &= 0.b^1_1b^1_2 \cdots b^1_{m} 1000 \cdots \\&\qquad\implies \alpha = 0.b^1_1 b^2_1 \cdots b^1_m b^2_m 1 b^2_{m+1} 0 b^2_{m+2} 0 b^2_{m+3} \cdots \\
&= 0.b^1_1b^1_2 \cdots b^1_{m} 0111 \cdots \\&\qquad\implies \alpha^\prime = 0.b^1_1 b^2_1 \cdots b^1_m b^2_m 0 b^2_{m+1} 1 b^2_{m+2} 1 b^2_{m+3} \cdots\\
\end{aligned}
$$

This shows the injectivity is broken, the same point $(v_1, v_2)$ can be mapped to by (at least) two different $\alpha \neq \alpha^\prime$. The Z-curve $Z^{-1}$ therefore no longer defines a one-to-one mapping, and is not injective.

For a point $(v_1, v_2) \in [0,1]^2$ which can be written as a finite binary expansion (that is, a $1$ followed by an infinite string of $0$'s), there might as much as 4 pre-images in $[0,1]$:

$$
\begin{aligned}
(v_1, v_2) &= (0.b^1_1 \cdots b^1_{m_1} 1000 \cdots, 0.b^2_1 \cdots b^2_{m_2} 1000 \cdots) \\
&= (0.b^1_1 \cdots b^1_{m_1} 1000 \cdots, 0.b^2_1 \cdots b^2_{m_2} 0111 \cdots) \\
&= (0.b^1_1 \cdots b^1_{m_1} 0111 \cdots, 0.b^2_1 \cdots b^2_{m_2} 1000 \cdots) \\
&= (0.b^1_1 \cdots b^1_{m_1} 0111 \cdots, 0.b^2_1 \cdots b^2_{m_2} 0111 \cdots) \\
\end{aligned}
$$

has the following pre-images (supposing without loss of generality $m_1 \leq m_2$):

$$
\begin{aligned}
\alpha_1 &= 0.b^1_1 b^2_1 \cdots b^1_{m_1} b^2_{m_1} 1 b^2_{m_1+1} 0 b^2_{m_1+2} 0 \cdots b^2_{m_2} 010000 \cdots \\
\alpha_2 &= 0.b^1_1 b^2_1 \cdots b^1_{m_1} b^2_{m_1} 1 b^2_{m_1+1} 0 b^2_{m_1+2} 0 \cdots b^2_{m_2} 010101 \cdots \\
\alpha_3 &= 0.b^1_1 b^2_1 \cdots b^1_{m_1} b^2_{m_1} 0 b^2_{m_1+1} 1 b^2_{m_1+2} 1 \cdots b^2_{m_2} 101010 \cdots \\
\alpha_4 &= 0.b^1_1 b^2_1 \cdots b^1_{m_1} b^2_{m_1} 0 b^2_{m_1+1} 1 b^2_{m_1+2} 1 \cdots b^2_{m_2} 101111 \cdots
\end{aligned}
$$

## The set of points with multiples preimages in $[0,1]$ is dense in $[0,1]^2$

The Z-curve is not injective from $[0,1]$ into $[0,1]^2$, and furthermore, we cannot ignore the points for which it is not injective, because the set of points with multiple preimages in $[0,1]$ is dense in $[0,1]^2$. We begin the study by considering only the first coordinate $v_1$, since the same arguments apply to $v_2$.

The set $\P$ is defined as the set of $v_1$ which have multiple pre-images in $[0,1]$:

$$
\P = \{ 0.b_1 b_2 \cdots b_m 1000 \cdots \ |\ m \in \mathbb{N} \}.
$$ {#eq-P}

We wish to prove that $\P$ is dense in $[0,1]$. For this, we will prove that any point $v_1 \in [0,1]$ is either in $\P$, or is the limit point of $\P$. If $v_1 \in \P$, we are done. Else, we consider $v_1 \notin \P$.

We define an expansion of $v_1 = \sum_{l=1}^\infty b_l 2^{-l}$ (we note $b_l$ are not unique, but this does not matter for the rest of this proof). We also define the truncation $v_1^{(m)} = \sum_{l=1}^m b_l 2^{-l}$, such that $v_1^{(m)} \xrightarrow{m\to\infty} v_1$. The trick is now to define the sequence of $v_1^\prime{^{(m)}}$ as:

$$
v_1^\prime{^{(m)}} = v_1^{(m)} + \frac{1}{2^{m+1}} \in \P.
$$

Therefore, the sequence $\{v_1^\prime{^{(m)}}\}$ is a subset of $\P$, and its limit is:

$$
\lim_{m\to\infty} v_1^\prime{^{(m)}} = \lim_{m\to\infty} v_1^{(m)} + \lim_{m\to\infty} \frac{1}{2^{m+1}} = v_1.
$$

Therefore, $v_1$ is a limit point of $\P$. This concludes the proof that $\P$ is dense in $[0,1]$.

By extension, the same arguments apply to $v_2$, and the set $\P \times \P$ is dense in $[0,1]\times [0,1]$. Therefore, the set of points with multiple preimages in $[0,1]$ is dense in $[0,1]^2$.

<!-- 
For this, we adopt a counting argument and look at the scaling behavior of $\P_n$, the finite-$n$ truncations of $\P$.

Let us take $n > m$, and the finite-$n$ truncations $v_1^{(n)}$ of $v_1 \in \P$. Because of the ambiguity of representation, we need to consider two different finite-$n$ truncations, that both converge to $v_1$ as $n\to\infty$. This defines the sequence of sets $\P_n$ as:

$$
\P_n = \{ 0.\underbrace{b_1 b_2 \cdots b_m}_{m\ \text{bits}} \underbrace{1000 \cdots 0}_{n-m\ \text{bits}} \ |\ m < n \} \cup \{b_1 b_2 \cdots b_m 0111\cdots 1\ |\ m < n\}.
$$ {#eq-Pn}

At finite $n$, we can define the complement, that is, the set of non-problematic points as:

$$
\begin{aligned}
\P_n^c &= \B_n \backslash \P_n \\
\text{where}\ \B_n &= \left\{ \frac{k}{2^n} \ |\ k \in \{0, \cdots, 2^{n}-1\}\right\}
\end{aligned}
$$ {#eq-Pnc}

We now argue based on the size of $\P$ and $\P^c$. From its definition, we directly see that $|\B_n| = 2^n$. To compute $|\P_n|$, it is easier to rewrite $\P_n$ as the union of $m$ *disjoint* sets $\P_{n,m}$:

$$
\begin{aligned}
\P_n &= \bigcup_{m=1}^n \P_{n,m} \\
&= \bigcup_{m=1}^n \left\{ \frac{k}{2^m} + \frac{1}{2^{m+1}}, \frac{k}{2^m} + \frac{2^{m+1}-1}{2^{m+1}}\ |\ k \in \{0, \cdots, 2^{m} - 1\},  \right\}.
\end{aligned}
$$

Since each set $\P_{n,m}$ is of size $|\P_{n,m}| = 2 \times 2^{m}$, the size of $|\P_n|$ can be computed as:

$$
|\P_n| = \sum_{m=1}^n |\P_{n,m}| = 2 \sum_{m=1}^n 2^m = 2 (2^n - 1)
$$

uh oh... -->
<!-- 
## $Z^{-1}$ defines a space-filling curve on a subset of $[0,1]$ {#sec-zcurve-lebesgue}

@@

talk about the fact we have other space filling curves, such as hilbert curve or peano curve, that are continuous and surjective [0,1] -> [0,1]^2. we use the Z-curve because it's easy to implement numerically -->

### 2023-06-16 manuscript/chapters/3_mappings.qmd
dkfjwen fhjk !!!!!

Peano functions seem like good candidates for the mapping $S^{-1}$, and are extensively studied in the literature around space-filling curves (we refer for instance to [@sagan_space-filling_1994]). They have the nice property that they are continuous and surjective, so we the segment populations can "reach" the entirety of the square populations. All things considered, non-injectivity is not a problem, since it just means that two segment populations can "refer" to the same square population, which does not break the regularity of the kernel, since the two segment populations can just have their own neighborhoods. Furthermore, we will see that in our examples, the non-injectivity can be ignored because the mapping is "quasi-bijective".

![Non-injectivity of $S^{-1}$ means that two segment populations can refer to the same square population.](figures/photo_2023-06-12_00-46-26.jpg)

@@

However, we argue that in our case only the notion of locality from 1D to 2D matters. In essence, the dynamics in the 2D embedding are already known, and we ask the question of whether it is possible to write an equivalent neural field 1D (in the sense that it has identical latent trajectories). Without repeating the intuition given in @sec-intro-simnf, if neighboring populations in 1D have similar potentials, then this allows us to write a neural field equation in $[0,1]$. Therefore we would like to have the property that populations close in 1D are (inversely) mapped to populations close in 2D, which already have similar potentials.

@@

@fig-anti-z gives some insight into the problem of this mapping. For a fixed input $(v_1,v_2)$, we see that the output point position jumps around when $n$ is changed, and that these jumps seem to always be of order 1. The intuition behind pointwise convergence is that as $n$ increases, the image of a point by $S^n$ is continually refined.

@@

The Z-mapping seems to converge to a fractal curve. Numerically, there seems to be an infinite number of discontinuities, but the size of these discontinuities is overwhelmingly small, so that *on average* they vanish as $n\to\infty$. We won't go further into detail about this in this section (in @sec-zcurve-lebesgue we develop the convergence to a space-filling fractal curve), but this gives some intuition for the arguments presented in @sec-locality.

@@

[^zmapping-notbijective]: Despite what can often be found in online resources, the Z-mapping is *not* a bijection because that would violate Netto's theorem. The reason for this is the ambiguity in the writing of binary expansions, and in @sec-zcurve-noninjective we give proof that the Z-mapping is not injective in the $n\to\infty$ limit. However, the Z-mapping is *quasi-bijective*, in the sense that the set of points for which the Z-mapping is not injective has zero measure, and can therefore be ignored.

>>>>

Peano functions seem like good candidates for the mapping $S^{-1}$, and are extensively studied in the literature around space-filling curves (we refer for instance to [@sagan_space-filling_1994]). They have the nice property that they are continuous and surjective, so we the segment populations can "reach" the entirety of the square populations. All things considered, non-injectivity is not a problem, since it just means that two segment populations can "refer" to the same square population (see @fig-non-injection), which does not break the regularity of the kernel, since the two segment populations can just have their own neighbourhoods.

![Non-injectivity of $S^{-1}$ means that two segment populations can refer to the same square population.](figures/photo_2023-06-12_00-46-26.jpg){#fig-non-injection}

@@

However, we argue that in our case only the notion of locality from 1D to 2D matters. In essence, the dynamics in the 2D embedding are already known, and we ask the question of whether it is possible to write an equivalent neural field 1D (in the sense that it has identical latent trajectories). Without repeating the intuition given in @sec-intro-simnf, if neighbouring populations in 1D have similar potentials, then this allows us to write a neural field equation in $[0,1]$. Therefore we would like to have the property that populations close in 1D are (inversely) mapped to populations close in 2D, which already have similar potentials.

@@

@fig-anti-z gives some insight into the problem of this mapping. For a fixed input $(v_1,v_2)$, we see that the output point position jumps around when $n$ changes, and that these jumps seem to always be of order 1. The intuition behind pointwise convergence is that as $n$ increases, the image of a point by $S^n$ is continually refined.

@@

The Z-mapping seems to converge to a fractal curve. Numerically, there seems to be an infinite number of discontinuities, but the size of these discontinuities is overwhelmingly small, so that *on average* they vanish as $n\to\infty$. We won't go further into detail about this (we refer to @Sagan1994 for an in-depth study), but this gives some intuition for the arguments presented in @sec-locality.

@@

[^zmapping-notbijective]: Despite what can often be found in online resources, the Z-mapping is *not* a bijection because that would violate Netto's theorem. For every point in $[0,1]^2$, there are at most 4 distinct preimages in $[0,1]$. The reason for this is the ambiguity in the writing of binary expansions, and in @sec-zcurve-noninjective we give proof that the Z-mapping is not injective in the $n\to\infty$ limit.
